A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-12-07 inst/manuscript/manuscript.tex
section 3,4,5 updates

FBPK assumes \(\bm{\delta}\) in Equation\(~\)\ref{equation:Zmarginal}
has mean zero and a spatial correlation structure that can be modeled
using a covariance function. This covariance function is commonly
assumed to be non-negative (between zero and one), second-order
stationary (depending only on the distance between population units),
isotropic (independent of direction), and decay with distance between
population units (Cressie, 1993). Henceforth, it is implied that we have
made these same assumptions regarding \(\bm{\delta}\), though Chiles and
Delfiner (1999), pp.~80-93 discuss covariance functions that are not
second-order stationary, not isotropic, or both. A variety of flexible
covariance functions can be used to model \(\bm{\delta}\) (Cressie,
1993); one example is the exponential covariance function (for a
thorough list of spatial covariance functions, see Cressie (1993). The
\(i,j\)th element of the exponential covariance matrix,
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\), is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is the variance parameter quantifying
the variability that is dependent (coarse-scale), \(\sigma^2_{2}\) is
the variance parameter quantifying the variability that is independent
(fine-scale), \(\phi\) is the range parameter measuring the
distance-decay rate of the covariance, and \(h_{i,j}\) is the Euclidean
distance between population units \(i\) and \(j\). The proportion of
variability attributable to dependent random error is
\(\sigma^2_{1} / (\sigma^2_{1} + \sigma^2_{2})\). Similarly, the
proportion of variability attributable to independent random error is
\(\sigma^2_{2} / (\sigma^2_{1} + \sigma^2_{2})\). Finally we note that
\(\sigma^2_{1}\) and \(\sigma^2_{2}\) are often called the partial sill
and nugget, respectively.

@@

\hypertarget{sec:numstudy}{%
\section{Numerical Study}\label{sec:numstudy}}

We used a simulation study to investigate performance of four
sampling-analysis combinations: IRS-Design, IRS with a design-based
analysis; IRS-Model, IRS with a model-based analysis; GRTS-Design, GRTS
sampling with a design-based analysis; and GRTS-Model, GRTS sampling
with a model-based analysis. These combinations are also provided in
Table \ref{tab:designanalysis}.

@@

Performance of the four sampling-analysis combinations was evaluated in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts, two response
types, and three proportions of dependent random error. The three sample
sizes (\(n\)) were \(n = 50, n = 100,\) and \(n = 200\). Samples were
always selected from a population size (\(N\)) of \(N = 900\). The two
location layouts were random and gridded. Locations in the random layout
were selected randomly from the unit square (\([0, 1] \times [0, 1]\)).
Locations in the gridded layout were selected randomly on a fixed grid
from the unit square. The two response types were normal and lognormal.
For the normal response type, the response was simulated using mean-zero
random errors with the exponential covariance
(Equation\(~\)\ref{equation:expcov}) for varying proportions of
dependent random error. The proportion of dependent random error is
represented by \(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where
\(\sigma^2_1\) and \(\sigma^2_2\) are from
Equation\(~\)\ref{equation:expcov}. The total variance,
\(\sigma^2_1 + \sigma^2_2\), was always 2. The range was always
\(\sqrt{2} / 3\), which means that the correlation in the dependent
random error decayed to nearly zero at the largest possible distance
between two units in the domain. For the lognormal response type, the
response was first simulated using the same approach as for the normal
response type, except that the total variance was 0.6931 instead of 2.
The response was then exponentiated, yielding a random variable whose
total variance is 2. The lognormal responses were used to evaluate
performance of the sampling-analysis approaches for data that were
skewed.

@@

In each of the 36 simulation scenarios, there were 2000 independent
simulation trials. In each trial, IRS and GRTS samples were selected and
then design-based and model-based analyses were used to estimate the
mean and construct confidence (design-based) or prediction (model-based)
intervals. We recorded the bias, squared error, and interval coverage
for all sampling-analysis combinations in each trial. Then we summarized
the performance of the combinations across trials by calculating average
bias, rMS(P)E (root-mean-squared error for the design-based approaches
and root-mean-squared-prediction error for the model-based approaches),
and the rate at which the true mean is contained in its 95\% interval.
The GRTS algorithm and the local neighborhood variance estimator are
available in the \textbf{\textsf{R}} package \texttt{spsurvey} (Dumelle
et al., 2021). FPBK is available in the \texttt{sptotal}
\textbf{\textsf{R}} package (Higham et al., 2021) and covariance
parameters were estimated using Restricted Maximum Likelihood (Harville,
1977; Patterson and Thompson, 1971; Wolfinger et al., 1994).

@@

Figure \ref{fig:figeff} shows the relative rMS(P)E of the four
approaches from Table \ref{tab:designanalysis} using the random location
layout with ``IRS-Design'' as the baseline. More formally, the relative
rMS(P)E is defined as \begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*} When there is no spatial correlation (Figure
\ref{fig:figeff}, top row), the four sampling-analysis combinations have
approximately equal rMS(P)E. So, using GRTS or using a spatial model
does not result in much, if any, loss in efficiency even when the
response variable is not spatially correlated. When there is spatial
correlation (Figure \ref{fig:figeff}, middle and bottom row), the
GRTS-Model combination tends to perform best, followed by GRTS-Design,
IRS-Model, and finally IRS-Design, though the difference in relative
rMS(P)E among IRS-Model, GRTS-Design, and GRTS-Model is relatively
small. As the strength of spatial correlation increases, the gap in
rMS(P)E between IRS-Design and the other combinations widens. Finally we
note that when there is spatial correlation, IRS-Model outperforms
IRS-Design by a large margin, suggesting that the poor design properties
of IRS are largely mitigated by the model-based analysis. These
conclusions are similar to those observed in the grid location layout.
Tables for rMS(P)E in all 36 simulation scenarios are provided in the
supplementary material.

@@

We also studied 95\% interval coverage among the combinations. The
design-based 95\% confidence intervals and model-based 95\% prediction
intervals were constructed using the normal distribution. Justification
for the design-based and model-based intervals comes from the asymptotic
normality of totals via the Central Limit Theorem.

Figure \ref{fig:figconf} shows the 95\% interval coverage for each of
the four combinations in the random location layout. All four
combinations have fairly similar interval coverage within each scenario.
Coverage in the normal response scenarios tended to be near 95\% and
slightly higher than coverage in the lognormal scenarios. Coverage in
the lognormal scenarios still generally exceeded 90\%. Coverage tended
to always increase with the sample size. At a sample size of 200, all
four combinations had approximately 95\% interval coverage in both
response scenarios and all dependent error proportions. These
conclusions were similar to those found in the grid location layout.
Tables for interval coverage in all 36 simulation scenarios are provided
in the supplementary material.

@@

We selected a single IRS sample and a single GRTS sample and estimated
the mean mercury concentration and its standard error using using
design-based and model-based approaches; Table \ref{tab:appliedtab}
shows the results. For all four sampling-analysis combinations, the true
realized mean mercury concentration is within the bounds of the 95\%
intervals. However, we should not generalize these results to any other
data or even to other samples from these data. But, we do note a couple
of patterns. The design-based IRS analysis shows the largest standard
error: a likely reason is that this is the only approach that does not
incorporate any spatial information regarding mercury concentration
across the contiguous United States. We also see that both approaches
using the GRTS sample have a lower standard error than the both
approaches using the IRS sample. We would expect this to be the case for
most samples because mercury concentration exhibits spatial patterning,
so a spatially balanced sample should usually yield a lower standard
error.

To better understand the dependence structure in mercury concentration,
the empirical semivariogram and corresponding fit of the model-based
approaches can be visualized. The empirical semivariogram quantifies the
halved squared differences (semivariance) among response values at
different distances apart. If a process exhibits strong spatial
dependence, the empirical semivariogram will have small values at small
distances and large values at large distances. Figure \ref{fig:figsv}
shows the empirical semivariogram for GRTS-Model, displaying the average
semivariance for several distances. Overlain onto Figure \ref{fig:figsv}
is the estimated semivariance obtained using the covariance parmaeters
from the REML fit of GRTS-Model. Figure \ref{fig:figsv} provides
evidence that there is strong correlation in mercury concentration among
the sites.

@@

The design-based and model-based approaches to inference are
fundamentally different paradigms by which samples are selected and data
are analyzed. The design-based approach incorporates randomness through
sampling to estimate a population parameter. The model-based approach
incorporates randomness through distributional assumptions to predict
the realized values of a random process. Though these approaches have
often been compared in the literature both from theoretical and
analytical perspectives, our contribution lies in studying them in a
spatial context while implementing spatially balanced sampling. Aside
from the theoretical differences described, a few analytical findings
from the simulation study are particularly notable. First, the sampling
decision (GRTS vs IRS) is most important when using a design-based
analysis. Though GRTS-Model still outperformed IRS-Model, the
model-based analysis mitigated much of the inefficiency of the IRS
sample. Second, independent of the analysis approach, there is no reason
to use IRS over GRTS for sampling spatial data, as GRTS-Design and
GRTS-Model generally performed at least as well as their IRS
counterparts when there was no spatial correlation and noticeably better
than there IRS counterparts when there was spatial correlation. Third,
The stronger the spatial correlation, the larger the gap in rMS(P)E
between IRS-Design and the other sampling-analysis combinations. Fourth
and finally, interval coverage for the normal response was very close to
95\% for all sample sizes, while interval coverage for the lognormal
response was not very close to 95\% until \(n = 200\).

There are several benefits and drawbacks of the design-based and
model-based approaches for spatial data, some of which we have not yet
discussed but are worthy of consideration in future research.
Design-based approaches are often computationally efficient, while
model-based estimation of covariance parameters can be computationally
burdensome, especially for likelihood-based methods such as REML that
rely on inverting a covariance matrix. The design-based approach also
more naturally handles binary data, free from the more complicated
logistic regression formulation commonly used to handle binary data in a
model-based approach. The model-based approach, however, can more
naturally quantify the relationship between covariates (predictor
variables) and the response variable. The model-based approach also
yields estimated spatial covariance parameters, which help better
understand the process of study. Model selection is also possible using
model-based approaches and criteria such as cross validation, likelihood
ratio tests, or AIC (Akaike, 1974). Model-based approaches are capable
of more efficient small-area estimation than design-based approaches by
leveraging distributional assumptions in areas with few observed sites.
Model-based approaches can also compute site-by-site predictions at
unobserved locations and use them to construct informative
visualizations. The benefits and drawbacks of both approaches, alongside
our theoretical and analytical comparisons, should be seriously
considered when choosing among them. This is especially true from an
analysis perspective, as we found that using a spatially balanced
sampling algorithm benefits both design-based and model-based analyses.

>>>>

FBPK assumes \(\bm{\delta}\) in Equation\(~\)\ref{equation:Zmarginal}
has mean-zero and a spatial correlation structure that can be modeled
using a covariance function. This covariance function is commonly
assumed to be non-negative (between zero and one), second-order
stationary (depending only on the distance between population units),
isotropic (independent of direction), and decay with distance between
population units (Cressie, 1993). Henceforth, it is implied that we have
made these same assumptions regarding \(\bm{\delta}\), though Chiles and
Delfiner (1999), pp.~80-93 discuss covariance functions that are not
second-order stationary, not isotropic, or both. A variety of flexible
covariance functions can be used to model \(\bm{\delta}\) (Cressie,
1993); one example is the exponential covariance function (for a
thorough list of spatial covariance functions, see Cressie (1993). The
\(i,j\)th element of the exponential covariance matrix,
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\), is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is the variance parameter quantifying
the variability that is dependent (coarse-scale), \(\sigma^2_{2}\) is
the variance parameter quantifying the variability that is independent
(fine-scale), \(\phi\) is the range parameter measuring the
distance-decay rate of the covariance, and \(h_{i,j}\) is the Euclidean
distance between population units \(i\) and \(j\). The proportion of
variability attributable to dependent random error is
\(\sigma^2_{1} / (\sigma^2_{1} + \sigma^2_{2})\). Similarly, the
proportion of variability attributable to independent random error is
\(\sigma^2_{2} / (\sigma^2_{1} + \sigma^2_{2})\). Finally we note that
\(\sigma^2_{1}\) and \(\sigma^2_{2}\) are often called the partial sill
and nugget, respectively.

@@

\hypertarget{sec:numstudy}{%
\section{Simulation Study}\label{sec:numstudy}}

We used a simulation study to investigate performance of four
sampling-analysis combinations: IRS with a design-based analysis, called
``IRS-Design''; IRS with a model-based analysis, called ``IRS-Model'';
GRTS sampling with a design-based analysis, called ``GRTS-Design''; GRTS
sampling with a model-based analysis, called ``GRTS-Model''. These
combinations are also provided in Table \ref{tab:designanalysis}.

@@

Performance for the four sampling-analysis combinations was evaluated in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts, two response
types, and three proportions of dependent random error. The three sample
sizes (\(n\)) were \(n = 50, n = 100,\) and \(n = 200\). Samples were
always selected from a population size (\(N\)) of \(N = 900\). The two
location layouts (of the population units) were random and gridded.
Locations in the random layout were randomly generated inside the unit
square (\([0, 1] \times [0, 1]\)). Locations in the gridded layout were
placed on a fixed, equall spaced grid inside the unit square. The two
response types were normal and lognormal. For the normal response type,
the response was simulated using mean-zero random errors with the
exponential covariance (Equation\(~\)\ref{equation:expcov}) for varying
proportions of dependent random error. The proportion of dependent
random error is represented by
\(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where \(\sigma^2_1\) and
\(\sigma^2_2\) are the dependent random error variance (partial sill)
and independent random error variance (nugget), respectively, from
Equation\(~\)\ref{equation:expcov}. The total variance,
\(\sigma^2_1 + \sigma^2_2\), was always 2. The range was always
\(\sqrt{2} / 3\), which means that the correlation in the dependent
random error decayed to nearly zero at the largest possible distance
between two units in the domain. For the lognormal response type, the
response was first simulated using the same approach as for the normal
response type, except that the total variance was 0.6931 instead of 2.
The response was then exponentiated, yielding a random variable whose
total variance is 2. The lognormal responses were used to evaluate
performance of the sampling-analysis approaches for data that were
skewed (i.e., not normal).

@@

In each of the 36 simulation scenarios, there were 2000 independent
simulation trials. In each trial, IRS and GRTS samples were selected and
then design-based and model-based analyses were used to estimate
(design-based) or predict (model-based) the mean and construct
confidence (design-based) or prediction (model-based) intervals. Then we
recorded the bias, squared error, and interval coverage for all
sampling-analysis combinations. After all 2000 trials, we summarized the
long-run performance of the combinations by calculating average bias,
rMS(P)E (root-mean-squared error for the design-based approaches and
root-mean-squared-prediction error for the model-based approaches), and
the proportion of tiems the true mean is contained in its 95\% interval.
The GRTS algorithm and the local neighborhood variance estimator are
available in the \textbf{\textsf{R}} package \texttt{spsurvey} (Dumelle
et al., 2021). FPBK is available in the \texttt{sptotal}
\textbf{\textsf{R}} package (Higham et al., 2021) and covariance
parameters were estimated using Restricted Maximum Likelihood (Harville,
1977; Patterson and Thompson, 1971; Wolfinger et al., 1994).

@@

Figure \ref{fig:figeff} shows the relative rMS(P)E of the four
approaches from Table \ref{tab:designanalysis} using the random location
layout with ``IRS-Design'' as the baseline. The relative rMS(P)E is
defined as \begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*} When there is no spatial correlation (Figure
\ref{fig:figeff}, ``Prop DE: 0'' row), the four sampling-analysis
combinations have approximately equal rMS(P)E. So using the GRTS
sampling plan or a model-based analysis does not result in much, if any,
loss in efficiency compared to IRS-Design when there is no spatial
correlation. When there is spatial correlation (Figure \ref{fig:figeff},
``Prop DE: 0.5'' and ``Prop DE: 0.9'' rows), GRTS-Model tends to perform
best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though
the difference in relative rMS(P)E among GRTS-Model, GRTS-Design, and
IRS-Model is relatively small. As the strength of spatial correlation
increases, the gap in rMS(P)E between IRS-Design and the other
sampling-analysis combinations widens. Finally we note that when there
is spatial correlation, IRS-Model outperforms IRS-Design by a large
margin, suggesting that the poor design properties of IRS are largely
mitigated by the model-based analysis. These conclusions are similar to
those observed in the grid location layout, so we omit a grid location
layout figure here. Tables for rMS(P)E in all 36 simulation scenarios
are provided in the supplementary material.

@@

We also studied 95\% interval coverage among the sampling-analysis
combinations. The design-based confidence intervals and model-based
prediction intervals were constructed using the normal distribution.
Justification for this comes from the asymptotic normality of means via
the Central Limit Theorem.

Figure \ref{fig:figconf} shows the 95\% interval coverage for each of
the four sampling-analysis combinations in the random location layout.
Within each scenario, the sampling-analysis combinations tend to have
fairly similar interval coverage. Coverage in the normal response
scenarios was usually near 95\%, while coverage in the lognormal
response scenarios varied from from 90\% to 95\%. Coverage tended to
always increase with the sample size. At a sample size of 200, all four
sampling-analysis combinations had approximately 95\% interval coverage
in both response scenarios for all dependent error proportions. These
conclusions are similar to those observed in the grid location layout,
so we omit a grid location layout figure here. Tables for interval
coverage in all 36 simulation scenarios are provided in the
supplementary material.

@@

We selected a single IRS sample and a single GRTS sample and estimated
(design-based) or predicted (model-based) the mean mercury concentration
and its standard error using using design-based and model-based
approaches. For the model-based analyses, the exponential covariance was
used. Table \ref{tab:appliedtab} shows the results from these analyses.
For all four sampling-analysis combinations, the true realized mean
mercury concentration is within the bounds of the 95\% confidence
(design-based) or prediction (model-based) intervals. Though we should
not generalize these results to other samples from these data, we do
note a couple of patterns. The design-based IRS analysis shows the
largest standard error: a likely reason is that this is the only
approach that does not incorporate any spatial information regarding
mercury concentration. Both analyses using GRTS sampling have lower
standard errors than both analyses using IRS sampling. We expect that
these patterns are consistent with other samples from these data because
mercury concentration exhibits spatial patterning, so a spatially
balanced sample should usually yield a lower standard error.

To better understand the dependence structure in mercury concentration,
the empirical semivariogram can be visualized alongside a fit using the
model-based approaches. The empirical semivariogram quantifies the
halved squared differences (semivariance) among response values at
different distances apart. If a process exhibits strong spatial
dependence, the empirical semivariogram will have small values at small
distances and large values at large distances. Figure \ref{fig:figsv}
shows the empirical semivariogram quantifying the average semivariance
for several distances. Overlain onto Figure \ref{fig:figsv} is the
estimated semivariance obtained using the covariance parmaeters from the
REML fit of GRTS-Model. Figure \ref{fig:figsv} provides evidence that
there is strong correlation in mercury concentration among the sites.

@@

The design-based and model-based approaches to statistical inference are
fundamentally different paradigms by which samples are selected and data
are analyzed. The design-based approach incorporates randomness through
sampling to estimate population parameters. The model-based approach
incorporates randomness through distributional assumptions to predict
realized values of a random process. Though these approaches have often
been compared in the literature both from theoretical and analytical
perspectives, our contribution lies in studying them in a spatial
context while implementing spatially balanced sampling. Aside from the
theoretical differences described, a few analytical findings from the
simulation study are particularly notable. First, the sampling decision
(GRTS vs IRS) is most important when using a design-based analysis.
Though GRTS-Model still outperformed IRS-Model, the model-based analysis
mitigated much of the inefficiency of the IRS sample. Second,
independent of the analysis approach, we found no reason to prefer IRS
over GRTS for sampling spatial data -- GRTS-Design and GRTS-Model
generally performed at least as well as their IRS counterparts when
there was no spatial correlation and noticeably better than their IRS
counterparts when there was spatial correlation. Third, as the strength
of spatial correlation increases, the gap in rMS(P)E between IRS-Design
and the other sampling-analysis combinations also increases. Fourth and
finally, when the response was normal, interval coverage for all
sampling-analysis combinations was very close to 95\% for all sample
sizes; when the response was lognormal, interval coverage for all
sampling and analysis was between 90\% and 95\% and closest to 95\% when
\(n = 200\).

There are several benefits and drawbacks of the design-based and
model-based approaches for finite population spatial data. Some we have
discuss, but others we have not and they are worthy of consideration in
future research. Design-based approaches are often computationally
efficient, while model-based estimation can be computationally
burdensome, especially for likelihood-based methods such as REML that
rely on inverting a covariance matrix. The design-based approach also
more naturally handles binary data, free from the more complicated
logistic regression framework commonly used to analyze binary data in a
model-based approach. The model-based approach, however, can more
naturally quantify the relationship between covariates (predictor
variables) and response variable. The model-based approach also yields
estimated spatial covariance parameters, which help better understand
the dependence structure in the process of study. Model selection is
also possible using model-based approaches and criteria such as cross
validation, likelihood ratio tests, or AIC (Akaike, 1974). Model-based
approaches are capable of more efficient small-area estimation than
design-based approaches by leveraging distributional assumptions in
areas with few observed sites. Model-based approaches can also compute
site-by-site predictions at unobserved locations and use them to
construct informative visualizations. The benefits and drawbacks of both
approaches, alongside our theoretical and analytical comparisons, can
motive the process of choosing among them. This is especially true from
an analysis perspective, as we found that using a spatially balanced
sampling algorithm benefits both design-based and model-based analyses.
